FAIR theory: Applying Open Science Principles to the Construction and Iterative Improvement of Scientific Theories

### 2025-01-08 manuscript.Rmd
update

Secondly, theory (meta)data should use vocabularies that follow FAIR principles (I2).
This is essentially a call to establish standardized ontologies,
which are themselves a type of theory [@meehlAppraisingAmendingTheories1990]. 
Thirdly, theory (meta)data should include qualified references to other (meta)data, including previous versions of the theory (I3).
The first part of this principle allows for nested theories;
for example, a theory that specifies causal relationships between constructs could refer back to an ontological theory from which those constructs are derived.
This can be achieved by linking the DOI of those nested theories [@ContributingCitationsReferences].
However, "[i]n systems with many dependencies, releasing new [...] versions can quickly become a nightmare."
The tension lies in balancing regular versioned updates with maintaining interconnected theories: changes risk breaking links between theories, but assuming every change does so leads to an overly sparse network.
In software development this problem is colloquially called "dependency hell" and usually addressed via semantic versioning.
Traditional version control only tracks changes but semantic versioning tracks what those changes mean to other users of the theory by using a specific format of version numbers.
We propose the following adaptation of semantic versioning for theories:

@@

### Making Theories FAIR Accelerates Scientific Progress
Adopting the FAIR principles for theories can address key challenges in the current research landscape, where theories often remain isolated and underutilized. By making theories findable, accessible, interoperable, and reusable, researchers can ensure that their work is grounded in a shared, transparent, and cumulative body of knowledge. This approach enhances scholarly communication, allowing for greater scrutiny, replication, and collaboration across disciplines, ultimately leading to faster, more reliable, and more impactful scientific progress.

@@

>>>>

Secondly, theory (meta)data should use vocabularies that follow FAIR principles (I2).
This is essentially a call to establish standardized ontologies,
which are themselves a type of theory [@meehlAppraisingAmendingTheories1990]. 
Thirdly, theory (meta)data should include qualified references to other (meta)data, including previous versions of the theory (I3).
The first part of this principle allows for nested theories;
for example, a theory that specifies causal relationships between constructs could refer back to an ontological theory from which those constructs are derived.
This can be achieved by linking the DOI of those nested theories [@ContributingCitationsReferences].
The second part of this principle allows for tracing the provenance of a theory; keeping track of its prior versions and other theories that inspired it.
This can be achieved by using Git for version control and Zenodo for archiving.
Git tracks the internal provenance of a theory repository; Zenodo is used to archive relations to external references (e.g., papers that influenced the theory, previous theories that inspired it, models based upon the theory).

@@

### Making Theories FAIR Accelerates Scientific Progress

Adopting the FAIR principles for theories can address key challenges in the current research landscape, where theories often remain isolated and underutilized. By making theories findable, accessible, interoperable, and reusable, researchers can ensure that their work is grounded in a shared, transparent, and cumulative body of knowledge. This approach enhances scholarly communication, allowing for greater scrutiny, replication, and collaboration across disciplines, ultimately leading to faster, more reliable, and more impactful scientific progress.

@@

## 6. Making Changes

<!-- CJ: Aaron, I moved your part here because I don't think it relates to interoperability. But perhaps there's another place where it fits better? -->
Aside from technical solutions to However, "[i]n systems with many dependencies, releasing new [...] versions can quickly become a nightmare."
The tension lies in balancing regular versioned updates with maintaining interconnected theories: changes risk breaking links between theories, but assuming every change does so leads to an overly sparse network.
In software development this problem is colloquially called "dependency hell" and usually addressed via semantic versioning.
Traditional version control only tracks changes but semantic versioning tracks what those changes mean to other users of the theory by using a specific format of version numbers.
We propose the following adaptation of semantic versioning for theories:
